plyranges: a grammar for transforming genomics data

### 2018-04-13 paper.Rmd
endomorphism section
context for consistency section

@@

This is possible because every 
function call is an endomorphism: when the input is GRanges object the output
will also be a GRanges object. One advantage of endomorphism is that it 
does not  require any additional learning of classes beyond GRanges and 
the _DataFrame_ classes. This strongly deviates from the design of 
the GRanges Bioconductor packages, where many methods return a new class
upon return. The Bioconductor design enables efficient computing as users
are exposed to low-level features of its API which _plyranges_ abstracts
away.

The combination of verbs we have defined above cover most operations that
can be performed in the original _IRanges_ and _GenomicRanges_ packages
without the user being exposed to new classes. In those packages to perform
most operations requires users to learn many classes and perform 
additional manipulations to return the results of their computation back to a
GRanges object.

@@

Using the design principles defined above every function in `plyranges`
performs a single action on GRanges objects. The `plyranges` DSL implments the 
core verbs from the `dplyr` package and implements a genomic relational algebra
for transforming GRanges objects (table x, y). Each verb preserves the 
semantics of GRanges object and works with derivatives of the GRanges class. 
Both of these aspects reduce the cognitive load on a new user since most 
manipulations can be performed with a vocabulary of several verbs, rather 
than having to memorise function names that are nouns.

@@

A caveat to constructing a compatible API with `dplyr` is that 
`plyranges` makes extensive use of non-standard evaluation in R
(achieved via the `rlang` package @R-rlang). Simply, this
means that computations are performed and evaluated in the context of the 
GRanges objects, which emphasises the interactive nature of the API. This has
the disadvantage that programming with `plyranges` becomes more difficult 
because a user needs to know about non-standard evaluation and user defined
functions can be difficult to debug.

Method chaining via the pipe operator can also be difficult to debug,
as there multiple points of failure

>>>>

* replacing arguments with functions
* standardised inputs for reading/writing genomics data as GRanges objects

@@

Most function calls in `plyranges` are endomorphisms: when the input 
is GRanges object the output will also be a GRanges object. The use of
endmoprhism in the `plyranges` DSL enables a user to predict the structure of 
the output of their computations and does not require them to learn any 
additional classes beyond GRanges and DataFrames.

This design pattern strongly deviates from the design of the OO interface in 
`GenomicRanges`, where many methods return a new class that is unfamiliar to 
the user upon return. These low-level classes enable efficient computing at the cost
of complexity and are abstracted away in `plyranges` interface.

@@

As a consequence of the design principles defined above every function in `plyranges`
performs a single action on GRanges objects. The `plyranges` DSL implments the 
core verbs from the `dplyr` package and implements a genomic relational algebra
for transforming GRanges objects (table x, y). Each verb preserves the 
semantics of GRanges object and works with derivatives of the GRanges class. 
Both of these aspects reduce the cognitive load on a new user since most 
manipulations can be performed with a vocabulary of several verbs, rather 
than having to memorise function names that are nouns.

@@

A caveat to constructing a compatible interface with `dplyr` is that 
`plyranges` makes extensive use of non-standard evaluation in R
(achieved via the `rlang` package @R-rlang). Simply, this
means that computations are performed and evaluated in the context of the 
GRanges objects, which emphasises the interactive nature of the interface.
This has the disadvantage that programming with `plyranges` becomes more difficult 
because a user needs to know about non-standard evaluation and user defined
functions can be difficult to debug.
